QPLIB: A Library of Quadratic Programming Instances

### 2017-01-23 QPLIB-1.0.tex
replace \bf comments by \todo
replace {\tt by \texttt{ and similar for \it and \bf
- i think \tt, \it, \bf, etc., are deprecated

On the other hand, (MI)QP\todo{wasn't QP supposed to already include the mixed-integer case? so why now again MIQP?} is, in some sense, a considerably more expressive class than (MI)LP. Quadratic expressions are found, either naturally or after appropriate reformulations, in very many optimization problems \cite{Kochenberger2014}. Table \ref{tbl:application} provides a certainly non-exhaustive collection of applications that lead to formulations with either quadratic constraints, or quadratic objective function, or both. Also, in general any continuous function can be approximated with arbitrary accuracy (over a compact set) by a polynomial of arbitrary degree; in turn, every polynomial can be broken down to a system of quadratic expressions. Hence, (MI)QP is, in some sense, roughly as expressive as the whole of (MI)NLP. Of course this is, in principle, true for (MI)LP as well, but at the cost of much larger and much more complex formulations ({\bf citations?}). Hence, for many applications QP may represent the ``sweet spot'' between the effectiveness, but lower expressive power, of (MI)LP and the higher expressive power, but much higher computational cost, of (MI)NLP.

@@

\multicolumn{3}{l}{{\bf Classical Problems that can be formulated as MIQP}} \\[1pt]

@@

\multicolumn{3}{l}{{\bf Computational chemistry \& Molecular biology}} \\[1pt]

@@

\multicolumn{3}{l}{{\bf Computational geometry}} \\[1pt]

@@

{\bf Finance} \\

@@

\multicolumn{3}{l}{{\bf Process Networks}} \\[1pt]

@@

\multicolumn{3}{l}{{\bf Robotics}} \\[1pt]

@@

\multicolumn{3}{l}{{\bf Telecommunications}} \\[1pt]

@@

\multicolumn{3}{l}{{\bf Energy}} \\[1pt]

>>>>

On the other hand, (MI)QP\todo{wasn't QP supposed to already include the mixed-integer case? so why now again MIQP?} is, in some sense, a considerably more expressive class than (MI)LP. Quadratic expressions are found, either naturally or after appropriate reformulations, in very many optimization problems \cite{Kochenberger2014}. Table \ref{tbl:application} provides a certainly non-exhaustive collection of applications that lead to formulations with either quadratic constraints, or quadratic objective function, or both. Also, in general any continuous function can be approximated with arbitrary accuracy (over a compact set) by a polynomial of arbitrary degree; in turn, every polynomial can be broken down to a system of quadratic expressions. Hence, (MI)QP is, in some sense, roughly as expressive as the whole of (MI)NLP. Of course this is, in principle, true for (MI)LP as well, but at the cost of much larger and much more complex formulations\todo{citations?}. Hence, for many applications QP may represent the ``sweet spot'' between the effectiveness, but lower expressive power, of (MI)LP and the higher expressive power, but much higher computational cost, of (MI)NLP.

@@

\multicolumn{3}{l}{\textbf{Classical Problems that can be formulated as MIQP}} \\[1pt]

@@

\multicolumn{3}{l}{\textbf{Computational chemistry \& Molecular biology}} \\[1pt]

@@

\multicolumn{3}{l}{\textbf{Computational geometry}} \\[1pt]

@@

\textbf{Finance} \\

@@

\multicolumn{3}{l}{\textbf{Process Networks}} \\[1pt]

@@

\multicolumn{3}{l}{\textbf{Robotics}} \\[1pt]

@@

\multicolumn{3}{l}{\textbf{Telecommunications}} \\[1pt]

@@

\multicolumn{3}{l}{\textbf{Energy}} \\[1pt]

### 2017-01-23 QPLIB-3.tex
replace {\tt by \texttt{ and similar for \it and \bf
- i think \tt, \it, \bf, etc., are deprecated

In addition to spontaneous contribution we analysed the other generic
libraries of instances available  on internet and containing QP
instances. Namely, the libraries from which we gathered instances are
%
\begin{itemize}
 \item the {\tt BARON} library
 \url{http://www.minlp.com/nlp-and-minlp-test-problems};
%
\item the {\tt CUTEst} library
 \url{https://ccpforge.cse.rl.ac.uk/gf/project/cutest/wiki};
%
\item the {\tt GAMS Performance} libraries
 \url{http://www.gamsworld.org/performance/performlib.htm};
%
\item the {\tt MacMINLP} library
 \url{https://wiki.mcs.anl.gov/leyffer/index.php/MacMINLP};
%
\item the {\tt Meszaros} library
 \url{http://www.doc.ic.ac.uk/~im/00README.QP};
%
\item the {\tt MINLP} library
 \url{http://www.gamsworld.org/minlp/minlplib.htm};
%
\item the {\tt POLIP} library
 \url{http://polip.zib.de/pipformat.php}.
\end{itemize}

@@

During the development of the library, a discussion ensued about what
the expected gols that we wanted to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item represent as much as possible all the different categories of QP
       problems;
 %
 \item gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item include for each of the categories a set of well-diversified
       instances;
 %
 \item obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve the aforementioned goals, we performed the following two
filters, applied in cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S~\ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in {\tt GAMS} (cf.~Table \ref{t:solvers})
       capable of solving it, whose number depends on the category of the
       instance under consideration. We then filtered according to a relative
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives\todo{we may want to be a tad more specific
       about what ``features'' exactly meant there}. Finally, in order to only
       keep computationally challenging instances we ran a general purpose solver
       able to tackle all QP categories with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded\todo{this might actually be best characterized as
       a third filter}.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $251$ discrete instances and
$116$ continuous instances.

>>>>

In addition to spontaneous contribution we analysed the other generic
libraries of instances available  on internet and containing QP
instances. Namely, the libraries from which we gathered instances are
%
\begin{itemize}
 \item the \texttt{BARON} library
 \url{http://www.minlp.com/nlp-and-minlp-test-problems};
%
\item the \texttt{CUTEst} library
 \url{https://ccpforge.cse.rl.ac.uk/gf/project/cutest/wiki};
%
\item the \texttt{GAMS Performance} libraries
 \url{http://www.gamsworld.org/performance/performlib.htm};
%
\item the \texttt{MacMINLP} library
 \url{https://wiki.mcs.anl.gov/leyffer/index.php/MacMINLP};
%
\item the \texttt{Meszaros} library
 \url{http://www.doc.ic.ac.uk/~im/00README.QP};
%
\item the \texttt{MINLP} library
 \url{http://www.gamsworld.org/minlp/minlplib.htm};
%
\item the \texttt{POLIP} library
 \url{http://polip.zib.de/pipformat.php}.
\end{itemize}

@@

During the development of the library, a discussion ensued about what
the expected gols that we wanted to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item represent as much as possible all the different categories of QP
       problems;
 %
 \item gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item include for each of the categories a set of well-diversified
       instances;
 %
 \item obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve the aforementioned goals, we performed the following two
filters, applied in cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S~\ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in \texttt{GAMS} (cf.~Table \ref{t:solvers})
       capable of solving it, whose number depends on the category of the
       instance under consideration. We then filtered according to a relative
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives\todo{we may want to be a tad more specific
       about what ``features'' exactly meant there}. Finally, in order to only
       keep computationally challenging instances we ran a general purpose solver
       able to tackle all QP categories with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded\todo{this might actually be best characterized as
       a third filter}.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $251$ discrete instances and
$116$ continuous instances.

### 2017-01-23 format.tex
unify referencing of ((sub)sub)sections, add other todos
replace {\tt by \texttt{ and similar for \it and \bf
- i think \tt, \it, \bf, etc., are deprecated

Following the notation of Section~\ref{sec:QPbasic}, a generic Quadratic Program (QP) can be written as follows:
%
\begin{align*}
 \min \;\;
 & x^\top Q^0 x + b^0 x + f \\
 %
 & c^i_l \leq x^\top Q^i x + b^i x \leq c^i_u & i \in \Mcm \\
 %
 & l_j \leq x_j \leq u_j & j \in \Mcn  \\
 %
 & x_j \in \mathbb{Z} & j \in \Mcz
\end{align*}

@@

The QPLIB format is defined in Table~\ref{tab-qplib-format}.
The data is in free format (blanks separate values), but must occur in
the order given here. Any blank lines, or lines starting with any of the
characters !, \% or \# are ignored. Each term in the first column of
Table~\ref{tab-qplib-format} denotes a required value. Any strings beyond
those required on a given line will be regarded as comments and ignored.
Real values may either by in decimal or exponential formats; for the latter,
the exponent may be preceded by either the character {\tt D} or {\tt E},
e.g. {\tt 12.56D+2} or {\tt 12.56E+2}.

@@

$j$\; ${\tt var\_name}_j$ & \lbox{index (integer) and name (character string)

@@

$i$\; ${\tt cons\_name}_i$ & \lbox{index  (integer) and name (character string) for each

@@

The problem type is represented by a character string as one of the following:
\begin{description}
\item For continuous problems (i.e., all variables are continuous):
\begin{description}[leftmargin=!,labelwidth=\widthof{\ttfamily SMILQPC}]
\item [LP]      a linear program,
\item [LPQC]    a linear program with quadratic constraints,
\item [BQP]     a bound-constrained quadratic program,
\item [QP]      a quadratic program, and
\item [QPQC]    a quadratic program with quadratic constraints.
\end{description}
\item For integer problems (i.e., all variables are integer valued):
\begin{description}[leftmargin=!,labelwidth=\widthof{\ttfamily SMILQPC}]
\item [ILP]     an integer linear program,
\item [ILPQC]   an integer linear program with quadratic constraints,
\item [IBQP]    an integer bound-constrained quadratic program,
\item [IQP]     an integer quadratic program, and
\item [IQPQC]   an integer quadratic program with quadratic constraints.
\end{description}
\item For mixed-integer problems (i.e., there is a mix of continuous and
integer variables):
\begin{description}[leftmargin=!,labelwidth=\widthof{\ttfamily SMILQPC}]
\item [MILP]    a mixed-integer linear program,
\item [MILPQC]  a mixed-integer linear program with quadratic constraints,
\item [MIBQP]   a mixed-integer bound-constrained quadratic program,
\item [MIQP]    a mixed-integer quadratic program, and
\item [MIQPQC]  a mixed-integer quadratic program with quadratic constraints.
\end{description}
\todo[inline]{\small Stefan: I agree that leaving out information on
numerical properties of the matrices and vectors (convexity, etc.) is good here.
Such properties may also depend on tolerances, etc.
As above, I would suggest to eliminate LPQC and rename QPQC into QCP.
Having a string for ``unspecified'' may be nice, too, or one just uses MIQCQP in this case.}
\end{description}

>>>>

Following the notation of \S~\ref{sec:QPbasic}, a generic Quadratic Program (QP) can be written as follows:
%
\begin{align*}
 \min \;\;
 & x^\top Q^0 x + b^0 x + f \\
 %
 & c^i_l \leq x^\top Q^i x + b^i x \leq c^i_u & i \in \Mcm \\
 %
 & l_j \leq x_j \leq u_j & j \in \Mcn  \\
 %
 & x_j \in \mathbb{Z} & j \in \Mcz
\end{align*}

@@

The QPLIB format is defined in Table~\ref{tab-qplib-format}.
The data is in free format (blanks separate values), but must occur in
the order given here. Any blank lines, or lines starting with any of the
characters !, \% or \# are ignored. Each term in the first column of
Table~\ref{tab-qplib-format} denotes a required value. Any strings beyond
those required on a given line will be regarded as comments and ignored.
Real values may either by in decimal or exponential formats; for the latter,
the exponent may be preceded by either the character \texttt{D} or \texttt{E},
e.g. \texttt{12.56D+2} or \texttt{12.56E+2}.

@@

$j$\; $\texttt{var\_name}_j$ & \lbox{index (integer) and name (character string)

@@

$i$\; $\texttt{cons\_name}_i$ & \lbox{index  (integer) and name (character string) for each

@@

The problem type is represented by a character string as one of the following:
\begin{description}
\item For continuous problems (i.e., all variables are continuous):
\begin{description}[leftmargin=!,labelwidth=\widthof{\texttt{SMILQPC}}]
\item [LP]      a linear program,
\item [LPQC]    a linear program with quadratic constraints,
\item [BQP]     a bound-constrained quadratic program,
\item [QP]      a quadratic program, and
\item [QPQC]    a quadratic program with quadratic constraints.
\end{description}
\item For integer problems (i.e., all variables are integer valued):
\begin{description}[leftmargin=!,labelwidth=\widthof{\texttt{SMILQPC}}]
\item [ILP]     an integer linear program,
\item [ILPQC]   an integer linear program with quadratic constraints,
\item [IBQP]    an integer bound-constrained quadratic program,
\item [IQP]     an integer quadratic program, and
\item [IQPQC]   an integer quadratic program with quadratic constraints.
\end{description}
\item For mixed-integer problems (i.e., there is a mix of continuous and
integer variables):
\begin{description}[leftmargin=!,labelwidth=\widthof{\texttt{SMILQPC}}]
\item [MILP]    a mixed-integer linear program,
\item [MILPQC]  a mixed-integer linear program with quadratic constraints,
\item [MIBQP]   a mixed-integer bound-constrained quadratic program,
\item [MIQP]    a mixed-integer quadratic program, and
\item [MIQPQC]  a mixed-integer quadratic program with quadratic constraints.
\end{description}
\todo[inline]{\small Stefan: I agree that leaving out information on
numerical properties of the matrices and vectors (convexity, etc.) is good here.
Such properties may also depend on tolerances, etc.
As above, I would suggest to eliminate LPQC and rename QPQC into QCP.
Having a string for ``unspecified'' may be nice, too, or one just uses MIQCQP in this case.}
\end{description}

### 2017-01-23 methods.tex
replace {\tt by \texttt{ and similar for \it and \bf
- i think \tt, \it, \bf, etc., are deprecated

Solution methods for QP can, following \cite{neumaier}, be broadly organized in three categories: {\it incomplete}, {\it asymptotically complete}, {\it complete}, and {\it rigorous}. Incomplete methods are only able to identify solutions, often locally optimal according to a suitable notion, and may even fail to find one even when one exists; in particular, they are typically not able to determine that an instance is empty. Asymptotically complete methods can find a globally optimal solution with probability one in infinite time, but again they cannot prove that a given instance is infeasible. Complete methods find an approximate globally optimal solution within a prescribed optimality tolerance within finite time, or prove that none such exists (but see Sect.~\ref{s:complete} below); they are often referred to as \emph{exact} methods in the computational optimization community. Finally, {\it rigorous} methods find globally optimal solutions within given tolerances even in the presence of rounding errors, except for ``near-degenerate cases''. Since none of the solvers we are using can be classified as rigorous, we limit ourselves to declaring solvers complete.

@@

Most local search methods for *C* are iterative in nature, and need a starting point $x^0$ as input. The ($k+1$)-st iterate is obtained as $x^{k+1} = x^k + \alpha_k d^k$, where $\alpha_k$ (a scalar) is the {\it step length}, and $d^k$ (a vector) is the {\it search direction}, e.g.~belonging to a tangent manifold and having a negative directional derivative at $x^k$, in order to improving optimality while reducing infeasibility. Alternatively, $x^{k+1} = x^k + d^k(\alpha_k)$ where $d^k(\cdot)$ is an arc satisfying $d^k(0) = 0$. Local search methods for *I* are also iterative, but since defining a useful notion of descent direction is harder in presence of integer variables, $d^k$ and $\alpha_k$ are usually computed in different ways, depending on the application at hand.

@@

\paragraph{Active set methods.}
%
At each iteration $k$ the algorithm forms the {\it active set} $\mathcal{A}^k$ containing the (indices of the) {\it active constraints}, i.e., all of the equality constraints as well as the inequality constraints that are satisfied at equality at the current iterate $x^k$. A subproblem, consisting of a minimization of a certain auxiliary objective function and including only active constraints, is then solved to identify a good search direction $d^k$. An appropriate step length $\alpha_k$ is then found by checking for the inactive constraints (since these must be satisfied too). If at $x^{k+1}$ some of the previously inactive constraints have become active, or vice versa, $\mathcal{A}^k$ is updated accordingly. This general scheme works because the step is chosen so that the objective decreases, and the active set cannot repeat.

@@

\paragraph{Global Adaptive Random Search (GARS).}
%
This is a modification of an algorithm called {\it pure random search}, which consists in sampling a random point $x'$ from a given compact set known to contain a global optimum, and then sampling a new candidate solution $y$ in a neighborhood of $x'$, setting $x' \leftarrow y$ if $y$ improves $x'$, and repeating as long as a termination condition is not satisfied. The adaptivity stems from changing the distribution for sampling $y$ at run-time, depending on the quality of the solutions identified by the method. Since this method only depends on sampling and function evaluation, it is usually fast. In the GARS solver of \lgo, it provides a useful starting point for a subsequent local search procedure. Asymptotic global convergence is attained by restarting the random search from different initial points $x'$.

@@

When the BB algorithm is allowed to select coordinate directions corresponding to continuous variables, it is called {\it spatial} BB (sBB). Branching on continuous (rather than integer or binary) variables becomes necessary in the presence of nonconvex nonlinearities, as it happens e.g.~in QCQ, since the quality of the bounds improves as the feasible set in the current node gets smaller.

>>>>

Solution methods for QP can, following \cite{neumaier}, be broadly organized in three categories: \emph{incomplete}, \emph{asymptotically complete}, \emph{complete}, and \emph{rigorous}. Incomplete methods are only able to identify solutions, often locally optimal according to a suitable notion, and may even fail to find one even when one exists; in particular, they are typically not able to determine that an instance is empty. Asymptotically complete methods can find a globally optimal solution with probability one in infinite time, but again they cannot prove that a given instance is infeasible. Complete methods find an approximate globally optimal solution within a prescribed optimality tolerance within finite time, or prove that none such exists (but see Sect.~\ref{s:complete} below); they are often referred to as \emph{exact} methods in the computational optimization community. Finally, rigorous methods find globally optimal solutions within given tolerances even in the presence of rounding errors, except for ``near-degenerate cases''. Since none of the solvers we are using can be classified as rigorous, we limit ourselves to declaring solvers complete.

@@

Most local search methods for *C* are iterative in nature, and need a starting point $x^0$ as input. The ($k+1$)-st iterate is obtained as $x^{k+1} = x^k + \alpha_k d^k$, where $\alpha_k$ (a scalar) is the \emph{step length}, and $d^k$ (a vector) is the \emph{search direction}, e.g.~belonging to a tangent manifold and having a negative directional derivative at $x^k$, in order to improving optimality while reducing infeasibility. Alternatively, $x^{k+1} = x^k + d^k(\alpha_k)$ where $d^k(\cdot)$ is an arc satisfying $d^k(0) = 0$. Local search methods for *I* are also iterative, but since defining a useful notion of descent direction is harder in presence of integer variables, $d^k$ and $\alpha_k$ are usually computed in different ways, depending on the application at hand.

@@

\paragraph{Active set methods.}
%
At each iteration $k$ the algorithm forms the \emph{active set} $\mathcal{A}^k$ containing the (indices of the) \emph{active constraints}, i.e., all of the equality constraints as well as the inequality constraints that are satisfied at equality at the current iterate $x^k$. A subproblem, consisting of a minimization of a certain auxiliary objective function and including only active constraints, is then solved to identify a good search direction $d^k$. An appropriate step length $\alpha_k$ is then found by checking for the inactive constraints (since these must be satisfied too). If at $x^{k+1}$ some of the previously inactive constraints have become active, or vice versa, $\mathcal{A}^k$ is updated accordingly. This general scheme works because the step is chosen so that the objective decreases, and the active set cannot repeat.

@@

\paragraph{Global Adaptive Random Search (GARS).}
%
This is a modification of an algorithm called \emph{pure random search}, which consists in sampling a random point $x'$ from a given compact set known to contain a global optimum, and then sampling a new candidate solution $y$ in a neighborhood of $x'$, setting $x' \leftarrow y$ if $y$ improves $x'$, and repeating as long as a termination condition is not satisfied. The adaptivity stems from changing the distribution for sampling $y$ at run-time, depending on the quality of the solutions identified by the method. Since this method only depends on sampling and function evaluation, it is usually fast. In the GARS solver of \lgo, it provides a useful starting point for a subsequent local search procedure. Asymptotic global convergence is attained by restarting the random search from different initial points $x'$.

@@

When the BB algorithm is allowed to select coordinate directions corresponding to continuous variables, it is called \emph{spatial} BB (sBB). Branching on continuous (rather than integer or binary) variables becomes necessary in the presence of nonconvex nonlinearities, as it happens e.g.~in QCQ, since the quality of the bounds improves as the feasible set in the current node gets smaller.
